Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
Fix it.
add merlinux
fix some things pointed out by toon

\numberofauthors{3}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\and
\alignauthor Maciej Fijalkowski\\
       \email{fijal@merlinux.eu}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

@@

We attempt to use the technique of Tracing JIT Compilers
\cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006,
mason_chang_efficient_2007} in the context
of the PyPy project, \ie on programs that are interpreters for some
dynamic languages, including Python.  Tracing JIT compilers can greatly
speed up programs that spend most of their time in loops in which they
take similar code paths.  However, applying an unmodified tracing JIT to
a program that is itself a bytecode interpreter results in very limited
or no speedup.
In this paper we show how to guide tracing JIT compilers to greatly
improve the speed of bytecode interpreters.  One crucial point is to
unroll the bytecode dispatch loop, based on two hints provided by the
implementer of the bytecode interpreter.  We evaluate our technique by
applying it to two PyPy interpreters: one is a small example, and the
other one is the full Python interpreter.

@@

Tracing JITs are an idea initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM
\cite{andreas_gal_trace-based_2009} and has been tried for Adobe's Tamarin
ActionScript VM \cite{chang_tracing_2009}.

@@

Typically, programs executed by a tracing VM go through various phases:
\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

At first, when the program starts, everything is interpreted.
The interpreter does a small amount of lightweight profiling to establish which loops
are run most frequently. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method looks for
loops in the user program.

@@

The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce efficient machine code by generating
code from the operations in it.  The machine code can then be executed immediately,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions and indirect or virtual
calls.  When generating the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit the machine code and continue the execution by falling
back to interpretation.\footnote{There are more complex mechanisms in place to
still produce more code for the cases of guard failures
\cite{andreas_gal_incremental_2006}, but they are independent of the issues
discussed in this paper.}

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}

The tracer interprets these functions in a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (\eg the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter given in Figure \ref{fig:tlr-basic} computing
the square of 10000000\footnote{The result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it.}
using the bytecode of Figure \ref{fig:square}. The results for various
configurations are as follows:

@@

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. However, constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This alone brings no
improvement over the previous case.

>>>>

\numberofauthors{4}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{University of Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\and
\alignauthor Maciej Fijalkowski\\
       \affaddr{merlinux GmbH}\\
       \email{fijal@merlinux.eu}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

@@

We attempt to use the technique of Tracing JIT Compilers
in the context
of the PyPy project, \ie on programs that are interpreters for some
dynamic languages, including Python.  Tracing JIT compilers can greatly
speed up programs that spend most of their time in loops in which they
take similar code paths.  However, applying an unmodified tracing JIT to
a program that is itself a bytecode interpreter results in very limited
or no speedup.
In this paper we show how to guide tracing JIT compilers to greatly
improve the speed of bytecode interpreters.  One crucial point is to
unroll the bytecode dispatch loop, based on two hints provided by the
implementer of the bytecode interpreter.  We evaluate our technique by
applying it to two PyPy interpreters: one is a small example, and the
other one is the full Python interpreter.

@@

Tracing optimizations were initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} to dynamically optimize
machine code at runtime. Its techniques were then successfully used to implement
a JIT compiler for a Java
VM \cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006}.
Subsequently these tracing JITs were discovered to be a
relatively simple way to implement JIT compilers for dynamic languages
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM
\cite{andreas_gal_trace-based_2009} and has been tried for Adobe's Tamarin
ActionScript VM \cite{chang_tracing_2009}.

@@

Typically tracing VMs go through various phases when they execute a program:
\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

At first, when the program starts, everything is interpreted.
The interpreter does a small amount of lightweight profiling to establish which loops
are run most frequently. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
is executed. Since loops need a backward jump somewhere, this method looks for
loops in the user program.

@@

The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results. Such a trace can be
used to generate efficient machine code. This generated machine code is
immediately executable, and can be used in the next iteration of the loop.

Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example at conditions and indirect or virtual
calls.  When generating the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit the machine code and continue the execution by falling
back to interpretation.\footnote{There are more complex mechanisms in place to
still produce extra code for the cases of guard failures
\cite{andreas_gal_incremental_2006}, but they are independent of the issues
discussed in this paper.}

@@

\begin{figure}
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

# corresponding trace:
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
\caption{A simple Python function and the recorded trace.}
\label{fig:simple-trace}
\end{figure}

Let's look at a small example. Take the (slightly contrived) RPython code in
Figure \ref{fig:simple-trace}.
The tracer interprets these functions in a bytecode format that is an encoding of
the intermediate representation of PyPy's translation tool\-chain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as in the lower half of Figure \ref{fig:simple-trace}.

The operations in this sequence are operations of the above-mentioned intermediate
representation (\eg the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter given in Figure \ref{fig:tlr-basic} computing
the square of 10000000 using the bytecode of Figure \ref{fig:square}.\footnote{The result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it.}
The results for various
configurations are as follows:

@@

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. However, constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This alone brings an
improvement over the previous case, but is still slower than pure
interpretation.
